---
title: "ET decay model"
author: "Francesco Giardina and Beni Stocker"
date: "5/19/2022"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggpubr)
```

## Linear model

Assume a 60-days dry down event with the following setup:

-   The initial water available to plants across the rooting zone, $S_0$ is 100 mm.
-   ET (termed $T$, since it's only transpiration here) is a linear function of the remaining water stored $S_t$ and is independent of VPD. $$
    T(t) = \alpha S(t)/S_0
    $$
-   The change in plant-available water storage is $\Delta S=T$.

This leads to an exponential decay of both $S$ and $D$ with time. We can set $\alpha = 0.01$ d $^{-1}$.

```{r}
s0 <- 100
alpha <- 1
ntsteps <- 500
s_t <- rep(NA, ntsteps)
t_t <- rep(NA, ntsteps)
s_t[1] <- s0 

for (idx in 1:(ntsteps-1)){
  t_t[idx] <- alpha * s_t[idx] / s0
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}

df_lin <- tibble( time = 1:ntsteps, soilm = s_t, transp = t_t ) 

df_lin %>% 
  ggplot(aes(time, soilm)) + 
  geom_line(size = 1)

df_lin %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, transp)) + 
  geom_line(size = 1)

df_lin %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(100-soilm, transp)) + 
  labs(x=expression(integral(ET))) +
  geom_line(size = 1)
```

## Effect of rooting zone water storage capacity

Let's compare the same relationship as above but with two different $S_0$.

```{r}
s0 <- 50
s_t[1] <- s0 
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- alpha * s_t[idx] / s0
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}


df_lin_deep <- df_lin %>% 
  mutate(fet = transp / transp[1]) %>% 
  mutate(rzwsc = "Deep", cwd = cumsum(transp))

df_lin_shallow <- tibble(time = 1:ntsteps, soilm= s_t, transp = t_t) %>% 
  mutate(fet = transp / transp[1]) %>% 
  mutate(rzwsc = "Shallow", cwd = cumsum(transp))

df_lin2 <- bind_rows(df_lin_deep, df_lin_shallow)

a <- df_lin2 %>% 
  ggplot(aes(x = time, y = soilm, color = rzwsc)) + 
  geom_line(size = 1) +
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14),  
        legend.text=element_text(size=14),
        legend.title=element_text(size=14)) +
  scale_color_manual( 
    values = c("#D81B60", "#BBCC33") # red, green 
  ) +
  labs(x = "Time (d)", y = expression(paste("Soil moisture (mm)"))) +
  guides(color=guide_legend(title="Root zone storage capacity"))
a

b <- df_lin2 %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, transp, color = rzwsc)) + 
  geom_line(size = 1) +
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14),  
        legend.text=element_text(size=14),
        legend.title=element_text(size=14)) +
  scale_color_manual( 
    values = c("#D81B60", "#BBCC33") # red, green 
  ) +
  labs(x = "Time (d)", y = expression(paste("ET (mm ", d^-1, ")"))) +
  guides(color=guide_legend(title="Root zone storage capacity"))
b

df_lin2 %>%
  tidyr::drop_na() %>% 
  ggplot(aes(x = cwd, y = transp, color = rzwsc)) +
  labs(x=expression(integral(ET))) +
  geom_line(size = 1)

c <- df_lin2 %>%
  tidyr::drop_na() %>% 
  ggplot(aes(x = cwd, y = fet, color = rzwsc)) +
  labs(x=expression(integral(ET))) +
  geom_line(size = 1) + 
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14),  
        legend.text=element_text(size=14),
        legend.title=element_text(size=14)) +
  scale_color_manual( 
    values = c("#D81B60", "#BBCC33") # red, green 
  ) +
  labs(x = "Cumulative water deficit (mm)", y = "fET (-)") +
  guides(color=guide_legend(title="Root zone storage capacity"))
c
```

The question is: How does VPD affect the ET decline?

## Effect of maximum conductance

Let's compare the same relationship as above but with two plant strategies.

```{r}
s0 <- 100
s_t[1] <- s0 
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- 1.5 * alpha * s_t[idx] / s0
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}


df_lin_conservative <- df_lin %>% 
  mutate(fet = transp / transp[1]) %>% 
  mutate(Strategy = "Conservative", cwd = cumsum(transp))

df_lin_exploit <- tibble(time = 1:ntsteps, soilm = s_t, transp = t_t) %>% 
  mutate(fet = transp / transp[1]) %>% 
  mutate(Strategy = "Exploitative", cwd = cumsum(transp))

df_lin3 <- bind_rows(df_lin_conservative, df_lin_exploit)

d<- df_lin3 %>% 
  ggplot(aes(x = time, y = soilm, color = Strategy)) + 
  geom_line(size = 1) +
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14),  
        legend.text=element_text(size=14),
        legend.title=element_text(size=14)) +
  scale_color_manual( 
    values = c("#0072B2", "#f6ae09") # blue-orange 
  ) +
  labs(x = "Time (d)", y = expression(paste("Soil moisture (mm)")))
d

e <- df_lin3 %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, transp, color = Strategy)) + 
  geom_line(size = 1) +
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14),  
        legend.text=element_text(size=14),
        legend.title=element_text(size=14)) +
  scale_color_manual( 
    values = c("#0072B2", "#f6ae09") # blue-orange 
  ) +
  labs(x = "Time (d)", y = expression(paste("ET (mm ", d^-1, ")")))
e

df_lin3 %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, fet, color = Strategy)) + 
  geom_line(size = 1)

f <- df_lin3 %>%
  tidyr::drop_na() %>% 
  ggplot(aes(x = cwd, y = fet, color = Strategy)) +
  labs(x=expression(integral(ET))) +
  geom_line(aes(size = Strategy, linetype = Strategy)) +
  scale_size_manual(values = c(1.5, 1)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  theme_classic() +
  theme(axis.text=element_text(size=12), 
        axis.title=element_text(size=14),  
        legend.text=element_text(size=14),
        legend.title=element_text(size=14)) +
  scale_color_manual( 
    values = c("#0072B2", "#f6ae09") # blue-orange 
  ) + 
  labs(x = "Cumulative water deficit (mm)", y = "fET (-)")
f

### COMBINE FIGURES
ggarrange(a, b, c,
          labels = c("a", "b", "c"),
          ncol = 3, nrow = 1,
          common.legend = TRUE, # one common legend
          legend="bottom" 
          ) 
# save plot
ggsave("Box_1_s0.png", path = "./", width = 12, height = 4)

ggarrange(d, e, f,
          labels = c("d", "e", "f"),
          ncol = 3, nrow = 1,
          common.legend = TRUE, # one common legend
          legend="bottom" 
          ) 
ggsave("Box_1_conductance.png", path = "./", width = 12, height = 4)

```
